---
title: "data-structure"
editor: visual
---

## Introduction

A data set is a collection of observations. These observations consist of one or more variables, or quantities that can change from observation to observation.

R has several data structures for storing data. We will focus on the data.frame type.

It's helpful to begin with one of the many data sets that are included with R. In particular, the iris data set is a great data set to begin with.

The iris data set consists of 150 observations with 5 variables. By typing iris in the command line, we will see the entrie data set.

For brevity, we can use the head() function to look at the first few rows of the data set:

```{r}
head(iris)
```

The first four variables represent the length and width measurements of each iris flower's sepal and petal, and the last variable lists the species of the iris flower.

## Extracting Values from a Data Frame

Sometimes we may just care about certain aspects of the data set. There are a few ways to get a specific row or a specfic column of the data set (or even a specific value!)

To access specific rows or columns, we can use square brackets after the name of the data frame.

```{r}
#| eval: false

iris["row number","column number"]
```

So, if we wanted to pick out the value in the third row and the fourth column, we would use:

```{r}
iris[3,4]
```

Note that this gives us the third value in the Petal.Width column (the fourth column)

If we want to get an entire row or an entire column, we would leave that corresponding portion blank. For the entire fifth row of the iris data set, we have:

```{r}
iris[5,]
```

For the entire second column, we can use:

```{r}
iris[,2]
```

We can also access columns using the dollar sign operator and then using the column name:

```{r}
iris$Sepal.Width
```

## Exercises

1. Consider using the mtcars data set in R. What are two different ways that we could get just the values in the wt (weight) variable?

2. Using the mtcars data set in R, how could we find the 18th value in the mpg variable?

3. How could I get just the first 3 rows of this data set?

